The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-26 article_final/main.tex
added section 5

\RequirePackage{graphicx}
 % The following packages will be automatically loaded:
 % amsmath, amssymb, natbib, graphicx, url, algorithm2e
 \usepackage{booktabs}
 %\usepackage{rotating}% for sideways figures and tables
\usepackage{longtable}% for long tables

@@

% Norms and operators
\newcommand{\norm}[1]{\lVert #1\rVert}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{2}}}
\DeclarePairedDelimiter{\normc}{\|}{\|_{\mathrm{C}}}
\DeclarePairedDelimiter{\normn}{\|}{\|_{*}}
\newcommand{\normkfk}[1]{\|#1\|_{\text{KF-}k}}

@@

\subsection{\texorpdfstring{$\normn{\mM^{k}}$ and Neon}{Neon}}
After considering $\normf{\mM^k}$ and $\norms{\mM^k}$, it is natural to look at the nuclear norm $\normn{\mM^k}$.
\begin{lemma}\label{lemma:neon_update}
	When $\norm{\cdot} = \normn{\cdot}$, \equationref{eq:our_update} turns into:

	\begin{equation}\label{eq:neon_update}
		\mX^{k+1} = \mX^{k} - \eta_k \vu_1 \vv_1^\top
	\end{equation}
\end{lemma}
\begin{proof}
Since $\norm{\cdot}^\dagger = \norms{\cdot}$, the goal is to reach $\norms{\mM^k} = \sigma_1$.

Let us note that $\Delta = \vu_1 \vv_1^\top$ delivers this value. Indeed, $\normn{\Delta}=1$ and by the trace property and orthogonality of the singular vectors,
\[
\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, \vu_1 \vv_1^\top> = \tr \diag({\sigma_1, 0, \dots, 0}) = \norms{\mM^k}\,,
\]
which completes the proof.
\end{proof}

@@

>>>>

\RequirePackage{graphicx}
 % The following packages will be automatically loaded:
 % amsmath, amssymb, natbib, graphicx, url, algorithm2e
 \usepackage{booktabs}
 %\usepackage{rotating}% for sideways figures and tables
\usepackage{longtable}% for long tables
\usepackage{arydshln}% for dashed lines in tables

@@

% Norms and operators
\newcommand{\norm}[1]{\lVert #1\rVert}
\DeclarePairedDelimiter{\normf}{\|}{\|_\mathrm{F}}
\DeclarePairedDelimiter{\norms}{\|}{\|_{\mathrm{2}}}
\DeclarePairedDelimiter{\normc}{\|}{\|_{\mathrm{C}}}
\DeclarePairedDelimiter{\normn}{\|}{\|_{*}}
\newcommand{\normkfk}[1]{\|#1\|_{\text{KF-}k}}
\DeclarePairedDelimiter{\normfstar}{\|}{\|_\mathrm{F*}}
\DeclarePairedDelimiter{\normftwo}{\|}{\|_\mathrm{F2}}

@@

\subsection{\texorpdfstring{$\normn{\mM^{k}}$ and Neon}{Neon}}
After considering $\normf{\mM^k}$ and $\norms{\mM^k}$, it is natural to look at the nuclear norm $\normn{\mM^k}$.
\begin{lemma}\label{lemma:neon_update}
	When $\norm{\cdot} = \normn{\cdot}$, \equationref{eq:our_update} becomes
	\begin{equation}\label{eq:neon_update}
		\mX^{k+1} = \mX^{k} - \eta_k \vu_1 \vv_1^\top\,.
	\end{equation}
\end{lemma}
\begin{proof}
Since $\norm{\cdot}^\dagger = \norms{\cdot}$, the goal is to reach $\norms{\mM^k} = \sigma_1$. Let us note that $\Delta = \vu_1 \vv_1^\top$ delivers this value. Indeed, $\normn{\Delta}=1$ and by the trace property and orthogonality of the singular vectors,
\[
\<\mM^k, \Delta> =\<\mU \mSigma \mV^\top, \vu_1 \vv_1^\top> = \tr \diag({\sigma_1, 0, \dots, 0}) = \norms{\mM^k}\,,
\]
which completes the proof.
\end{proof}

@@

\section{Conic Combination of LMO-algorithms is an LMO-algorithm}
\label{sec:conic_combo}

Simply applying norm dual to the Ky Fan $k$-norm however, is not enough to provide family of algorithms diverse enough for our cause. Next we consider linear combinations of algorithms from LMO framework, which happen to be LMO-algorithms too, as we show in succeeding paragraphs.

\subsection{General Case}
First, we need a well-known fact mentioned, for example, in \citet[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof here.

\begin{lemma}\label{lemma:dual_to_conv_comb}
	Let $\norm{\cdot}_{(1)},\dots,\norm{\cdot}_{(n)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha_1,\dots,\alpha_n$ be non-negative reals. Define
	\[
		\norm{\cdot} := \sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}.
	\]
	Then the dual unit ball of $\norm{\cdot}$ satisfies
	\[
		\cB_{\norm{\cdot}^\dagger}
		= \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}
	\]
	where $\sum$ denotes the Minkowski sum and $\cB_{\norm{\cdot}_{(i)}^\dagger}$ is the unit ball of the dual norm $\norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

\begin{proof}
	Let us first prove the lemma for the case $n = 2$. Denote $f(x) = \alpha_1 \norm{x}_{(1)}$ and $g(x) = \alpha_2 \norm{x}_{(2)}$, such that $\norm{x} = f(x) + g(x)$. Recall two standard facts:
	\begin{enumerate}
		\item For any norm $\norm{\cdot}$ and $\lambda>0$,
		      \[
			      (\lambda \norm{\cdot})^*(y) =
			      \sup_{x}\bigl( \langle y,x \rangle - \lambda \norm{x}\bigr)
			      = \delta_{\lambda \cB_{\norm{\cdot}^\dagger}}(y)\,,
		      \]
		      i.e., the indicator function of the scaled dual ball.
		\item The Fenchel conjugate of a sum satisfies
		      \[
			      (f+g)^*(y) = \inf_{u+v=y} \bigl(f^*(u) + g^*(v)\bigr)\,.
		      \]
	\end{enumerate}
	Applying these to $f$ and $g$, we have
	\[
		f^*(u) = \delta_{\alpha_1 \cB_{\norm{\cdot}_{(1)}^\dagger}}(u)\,,
		\quad
		g^*(v) = \delta_{\alpha_2 \cB_{\norm{\cdot}_{(2)}^\dagger}}(v)\,.
	\]
	Thus,
	\[
		\norm{\cdot}^*(y)
		= (f+g)^*(y)
		= \inf_{u+v=y}
		\bigl(
		\delta_{\alpha_1 \cB_{\norm{\cdot}_{(1)}^\dagger}}(u)
		+
		\delta_{\alpha_2 \cB_{\norm{\cdot}_{(2)}^\dagger}}(v)
		\bigr)
		=
		\delta_{\alpha_1 \cB_{\norm{\cdot}_{(1)}^\dagger}+\alpha_2 \cB_{\norm{\cdot}_{(2)}^\dagger}}(y).
	\]
	By definition, the conjugate of a norm is exactly the indicator of its dual unit ball:
	\[
	\norm{\cdot}^*(y) = \delta_{\cB_{\norm{\cdot}^\dagger}}(y)\,.
	\]
	Therefore, $\cB_{\norm{\cdot}^\dagger} = \alpha_1 \cB_{\norm{\cdot}_{(1)}^\dagger} + \alpha_2 \cB_{\norm{\cdot}_{(2)}^\dagger}$.

	Now we prove the general case by induction. The base case ($n=2$) is already proven. Suppose that the assumption of the lemma holds for $n = k$. Then, for $n = k + 1$,
	\[
	\norm{x} = \sum_{i = 1}^k \alpha_i \norm{x}_{(i)} + \alpha_{k + 1}\norm{x}_{(k + 1)} = \norm{x}_{(1:k)} + \alpha_{k + 1}\norm{x}_{(k + 1)}\,.
	\]
	Applying the result for $n=2$ combined with the induction assumption, we obtain
	\[
	\cB_{\norm{\cdot}^\dagger} = \cB_{\norm{\cdot}_{(1:k)}^\dagger} + \alpha_{k + 1} \cB_{\norm{\cdot}_{(k + 1)}^\dagger} = \sum_{i = 1}^{k + 1} \alpha_i \cB_{\norm{\cdot}_{(i)}^\dagger}\,,
	\]
	which proves the lemma.
\end{proof}

\begin{lemma}\label{lemma:lin_comb_lmo}
	Let $\norm{\cdot}_{(1)}, \dots, \norm{\cdot}_{(n)}$ be norms on a finite-dimensional Euclidean space, and let $\alpha_1, \dots, \alpha_n$ be non-negative reals. Consider Linear Minimization Oracles $\mathrm{LMO}_{1}, \dots, \mathrm{LMO}_{n}$, corresponding to the unit balls of these norms. Then, $\sum_{i = 1}^n \alpha_i \mathrm{LMO}_{i}$ is the LMO corresponding to the norm $\norm{\cdot}$ dual to the $\sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}^\dagger$.
\end{lemma}

\begin{proof}
	Using \lemmaref{lemma:dual_to_conv_comb} and the fact $\norm{\cdot}^{\dagger \dagger} = \norm{\cdot}$, we can obtain general form of the unit ball in the $\norm{\cdot}$ norm: $\cB_{\norm{\cdot}} = \sum_{i = 1}^n \alpha_i \cB_{\norm{\cdot}_{(i)}}$. Thus, the linear function minimization objective on a $\norm{\cdot}$ norm ball can be transformed as follows:
	\[
	\argmin_{\mD \in \cB_{\norm{\cdot}}}\<\mM, \mD> = \argmin_{\mD_1 \in \alpha_i \cB_{\norm{\cdot}_{(1)}},\dots,\mD_n \in \alpha_n\cB_{\norm{\cdot}_{(n)}}}\<\mM, \sum_{i = 1}^n\mD_i> = \sum_{i = 1}^n \argmin_{D_i \in \cB_{\norm{\cdot}_{(i)}}}\<\mM, \mD_i>\,,
	\]
	where the last summation denotes the Minkowski sum. This immediately implies $\sum_{i = 1}^n \alpha_i \mathrm{LMO}_{i} \in \argmin_{\mD \in \cB_{\norm{\cdot}}}\<\mM, \mD>$, which proves the claim of the lemma.
\end{proof}

Applying it to optimization algorithms, we obtain the following corollary.

\begin{corollary}\label{corollary:lin_comb_alg}
	Let there be a finite family of LMO based algorithms indexed by $i = 1,\dots,n$, where the update of the $i$-th algorithm is defined by
	\[
	\mX^{k + 1} - \mX^{k} = \gamma_k \mathrm{LMO}_{i}(\mM^{k})\,,
	\]
	and $\mathrm{LMO}_i$ corresponds to the unit ball of norm $\norm{\cdot}_i$. For arbitrary non-negative $\alpha_1,\dots,\alpha_n$, the algorithm with the update given by
	\[
	\mX^{k + 1} - \mX^{k} = \gamma_k \sum_{i = 1}^n \alpha_i \mathrm{LMO}_{\norm{\cdot}_{(i)}}(\mM^k)
	\]
	is an LMO-algorithm itself, with LMO corresponding to the unit ball of the norm $\norm{\cdot}$ dual to the norm given by $\sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}^\dagger$.
\end{corollary}

\subsection{Frobeniusize Them: F-Muon and F-Neon}
Let us construct the concrete examples of algorithms given by linear combinations of LMO-algorithms. It follows from \corollaryref{corollary:lin_comb_alg} that those algorithms can also be viewed as LMO-algorithms.

Combining Fanions for various $k$ with another excellent LMO-algorithm, NSGD, we obtain the family of algorithms with updates defined by
\begin{equation}\label{eq:nsgd_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_k \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\frac{\mM^k}{\normf{\mM^k}}\right)\,.
\end{equation}
Recall that Fanion-$k$ update is governed by the norm dual to the Ky Fan-$k$ norm, and NSGD corresponds to the Frobenius norm, which is dual to itself. Thus, by \corollaryref{corollary:lin_comb_alg}, our new algorithm is an LMO-algorithm corresponding to the unit ball of the norm $\norm{\cdot}_{\mathrm{F-KF-k}}$:
\begin{equation}\label{eq:fkfk_norm}
	\norm{\cdot}_{\mathrm{F-KF-k}}^\dagger = \alpha \normkfk{\cdot} + (1-\alpha)\normf{\cdot}\,.
\end{equation}
We name the derived family of algorithms \emph{F-Fanions}.

Two edge members of this family, with $k = 1$ and $k = \min\{m, n\}$ correspondingly, \emph{F-Neon} and \emph{F-Muon}, are of particular interest to us. A bit of information and visualizations related to the $\normfstar{\cdot} = \norm{\cdot}_{\mathrm{F-KF-1}}$ norm is presented in the appendix (see \equationref{eq:normfstardual}, \figureref{fig:fstardual}).

\subsection{Add SignSGD: S-Muon and S-Neon}

Another algorithm we consider is the SignSGD. Mixing its update with that of Fanion-$k$, we obtain the update of \emph{S-Fanion-$k$}:
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_k \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^k)\right)\,,
\end{equation}
where $\eta$ is the special SignSGD learning rate coefficient.

The norm $\norm{\cdot}_{\mathrm{C-KF-k}}$ which produces this update in the LMO framework is again given by \corollaryref{corollary:lin_comb_alg}:
\begin{equation}\label{eq:ckfk_norm}
	\norm{\cdot}_{\mathrm{C-KF-k}}^\dagger = \alpha \normkfk{\cdot} + \frac{1-\alpha}{\eta}\normc{\cdot}^\dagger\,.
\end{equation}

\section{Computing the Updates}\label{sec:matrix-side}
We employ the thick-restarted Lanczos method for the symmetric eigenvalue problem (thick-restart Lanczos, TRLan) to compute the low-rank updates of Fanions. We apply TRLan to either $\mM^{k\top}\mM^k$ or $\mM^k \mM^{k\top}$ (whichever matrix is smaller). We use the CuPy implementation of \texttt{cupy.sparse.linalg.svds} \citep{cupy_svds_ref} which internally relies on TRLan \citep{simonz1998thick}.

TRLan is specifically designed for efficiently approximating the largest singular values and corresponding singular vectors of very large matrices. The thick-restart strategy retains the most informative Ritz vectors across restarts, which dramatically accelerates convergence while keeping memory consumption moderate. TRLan is particularly attractive in our GPU setting because its dominant cost is a modest number of highly parallelizable matrix-vector multiplications (matvecs) and it avoids full reorthogonalization against the entire Krylov basis by using short recurrence relations combined with thick restarting.

The per-cycle complexity is $\mathcal{O}(mn^2 + n^2k + nk^2)$, where $m \geq n$ are the dimensions of the target matrix and $k$ is the size of the retained subspace (typically $k \ll n$).

In \tableref{tbl:matrix_methods}, we compare TRLan against randomized SVD (RSVD) and simple power iterations when computing the rank-$k$ update used in Fanion-$k$ and related algorithms. Experiments are performed on dense random matrices with i.i.d. $\mathcal{N}(0,1)$ entries and use CPU implementations for fair comparison.
We report:
\begin{itemize}
  \item $err_1$: relative error in the Frobenius norm of $\sum_i^k \vu_i \vv_i^T$,
  \item $err_2$: relative error in the Frobenius norm of $\sum_i^k \sigma_i \vu_i \vv_i^T$.
\end{itemize}

On $500\times500$ matrices, TRLan and RSVD require comparable wall-clock time, but TRLan delivers orders-of-magnitude lower error and far fewer matvecs. On larger $5000\times5000$ matrices the advantage becomes even more pronounced: TRLan is 3-4 times faster than RSVD while using $\sim$30 times fewer matvecs at comparable or better accuracy.

An interesting empirical observation is that RSVD tends to approximate the \emph{singular values} themselves reasonably well but the reconstructed low-rank matrix noticeably deviates from the truncated SVD, whereas TRLan provides an excellent approximation to the truncated SVD matrix itself (low $err_2$) at the cost of occasionally less accurate individual singular values. This makes TRLan the preferred choice for algorithms like Neon/Fanion-$k$ that only need the low-rank term $\sum \sigma_i \vu_i \vv_i^\top$, but less ideal for methods (e.g., Dion) that explicitly require accurate $\sigma_i$ for error feedback or step-size control.

A current practical limitation is the lack of a native PyTorch implementation of thick-restart Lanczos; existing PyTorch-based randomized SVD routines cannot match TRLan's accuracy/efficiency combination for the matrix reconstruction task.

For reference, \tableref{tbl:newton_schulz_reference} shows results for the Newton-Schulz polar decomposition iteration on the same matrices, $err_1$ is the relative error of $\mU\mV^T$ (29-30 iterations to converge, significantly higher matvec count than TRLan).

\begin{table}[ht]
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|}
	  \hline
	  Matrix sizes & $k$  & Method           & Time (s) & Matvecs & Iterations & $err_1$         & $err_2$       \\
	  \hline
	  $500 \times 500$      & 5  & Power Iterations           & 0.062   & 2005    & 200        & 9.2$\cdot 10^{-3}$ & 9.1$\cdot10^{-3}$\\
	  $500 \times 500$      & 5  & RSVD             & 0.017   & 1170    & 38         & 9.8$\cdot 10^{-3}$ & 9.6$\cdot10^{-3}$\\
	  $500 \times 500$      & 5  & TRLan            & 0.018   & 131     & 65         & 9.6$\cdot 10^{-5}$ & 9.4$\cdot10^{-5}$\\
	  \hdashline
	  $500 \times 500$      & 50 & Power Iterations           & 0.44    & 43750   & 437      & 9.9$\cdot 10^{-3}$ & 9.0$\cdot10^{-3}$\\
	  $500 \times 500$      & 50 & RSVD             & 0.61    & 6120    & 50      & 9.9$\cdot 10^{-3}$ & 9.1$\cdot10^{-3}$\\
	  $500 \times 500$      & 50 & TRLan            & 0.16    & 462     & 231      & 3.3$\cdot 10^{-7}$ & 3.0$\cdot10^{-7}$\\
	  \hdashline
	  $5000 \times 5000$    & 5  & Power Iterations           & 9.6     & 9065    & 906      & 8.6$\cdot 10^{-3}$ & 8.6$\cdot10^{-3}$\\
	  $5000 \times 5000$    & 5  & RSVD             & 2.1     & 5640    & 187      & 9.7$\cdot 10^{-3}$ & 9.7$\cdot10^{-3}$\\
	  $5000 \times 5000$    & 5  & TRLan            & 0.70    & 205     & 102      & 7.7$\cdot 10^{-3}$ & 7.7$\cdot10^{-3}$\\
	  \hline
	\end{tabular}
  \caption{Comparison of methods for computing rank-$k$ updates on dense random matrices (CPU, double precision). Lower is better in all columns.}
  \label{tbl:matrix_methods}
  \end{table}

  \begin{table}[ht]
  \centering
  \begin{tabular}{|c|c|c|c|c|}
  \hline
  Matrix size    & Time (s) & Matvecs & Iterations & $err_1$ \\
  \hline
  $500 \times 500$     & 0.041    & 27 000  & 27         & 4.8e-3  \\
  $5000 \times 5000$   & 26.4     & 290 000 & 29         & 6.5e-3  \\
  \hline
  \end{tabular}
  \caption{Newton–Schulz iteration on the same matrices (for reference).}
  \label{tbl:newton_schulz_reference}
  \end{table}
